 

### 2022-05-13 thesis/discussion.qmd
Change layout

| reporting item        | description                                                                                                           |
|-----------------------|-----------------------------------------------------------------------------------------------------------------------|
| Metabolic Cart        | State the exact device model and manufacturer                                                                         |
| Measurement Mode      | State the measurement mode (e.g., mixing chamber, breath-by-breath,...)                                               |
| Software              | State the name and version of the software used for data analysis                                                     |
| Preprocessing         | State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis |
| Processing Strategy   | State the exact data processing strategy used to determine the V̇O~2max~ (e.g., binned time averages)                  |
| Processing Parameters | State the parameters used for the processing strategy (e.g., length of averaging interval)                            |
| Rationale             | State the rationale for using the processing strategy (e.g., reference to recommendations)                            |

: Recommendations for reporting data processing strategies to determine V̇O~2max~. {#tbl-recommendation}

>>>>

```{r}
#| echo: false
#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake"
#| label: tbl-recommendation

rec_df <- data.frame(
  item = c("Metabolic Cart", "Measurement Mode", "Software", "Preprocessing", "Processing Strategy", "Processing Parameters", "Rationale"),
  desc = c(
    "State the exact device model and manufacturer",
    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…)",
    "State the name and version of the software used for data analysis",
    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis",
    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages)",
    "State the parameters used for the processing strategy (e.g., length of averaging interval)",
    "State the rationale for using the processing strategy (e.g., reference to recommendations)")
)

out <- kableExtra::kbl(
  rec_df,
  col.names = c("Reporting Item", "Description"),
  format = "latex",
  booktabs = TRUE
) |>
   kableExtra::column_spec(
    2, width = "11cm"
  )

knitr::asis_output(stringr::str_replace_all(out, "VO2max", "${\\\\dot{\\\\textsf{V}}\\\\textsf{O}_{\\\\textsf{2max}}}$"))
```

### 2022-05-13 thesis/introduction.qmd
Change layout

### Maximum Oxygen Uptake: Definition and Relevance

Performance in endurance sports is limited by the human physiology. Athletes need to supply energy to the contracting muscles for locomotion, a process that happens mainly via the oxidative phosphorylation. A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster [@holloszy1984]. On a whole-body level the highest activity of oxidative phosphorylation can be approximated by measuring the maximum oxygen uptake (V̇O~2max~).

V̇O~2max~ is defined as the highest rate a body can consume oxygen. Determined by measuring gas exchange data, it is one of the most common assessed exercise parameters in sports science and medicine. V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded the most relevant physiological parameter for predicting endurance performance [@bassett2000]. Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~

### Measuring the Maximum Oxygen Uptake

Researchers measure V̇O~2max~ during exercise tests to exhaustion. In a laboratory setting such tests commonly take place on treadmills or cycling ergometer. Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value. Therefore most exercise protocols consist of a continuous or stepwise, fast increase in load. @buchfuhrer1983 and @yoon2007 suggested an optimal duration of 8-12 or 8-10 minutes for testing V̇O~2max~. While this narrow limits have been challenged, considerably shorter or longer protocols are advised against [@midgley2008]. Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.

Early research in V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017]. Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001]. While mixing chambers are regarded as more reliable [@beijst2013], measuring breath-by-breath generates data with a higher temporal resolution [@roecker2005]. The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently.

### Peak vs. Maximum Oxygen Uptake

Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018]. To identify a 'true' maximum, researchers commonly evaluate a set of parameters measured during the ramp test --- the so called 'V̇O~2max~ criteria' [@howley1995]. If these criteria are (partly) not fulfilled, researchers are advised to speak of a peak oxygen uptake instead of V̇O~2max~ [@howley1995]. In this thesis I will not distinguish between peak and maximum oxygen uptake, as the criteria for V̇O~2max~ (e.g. the primary criterion of a plateau in oxygen uptake or the secondary criterion of the maximum respiratory quotient) do by themselves heavily rely on the data processing strategy used [@astorino2009]. I thus speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.

@@

Measured breath-by-breath data is noisy (see grey points in @fig-example). Both the biological variability of breathing patterns and the measurement error --- as well as irregular breaths such as coughs and swallowing --- lead to a highly fluctuating raw oxygen uptake. For interpretation the raw data requires some form of processing.

Different data processing strategies influence measured parameters of gas exchange. As soon as when the first automated systems for gas exchange measurement were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values. Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998; @astorino2009; @sell2021]. @midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, highlighting that no optimal strategy exists in terms of reliability, but that consistency of strategies is key when comparing outcomes. Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

![Raw breath by breath data processed by different strategies during the final minutes of a ramp test to exhaustion. Grey points display oxygen uptake from the single breaths. The moving averages were calculated over 30 breaths and seconds, respectively. The binned average was calculated over 30-second intervals with the mean values aligned to the center of each interval. For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied. For more details on the used data processing strategies, see the methods section.](../plots/example.png){#fig-example}

@@

Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data, @robergs2003a opposed these conclusion, stating that shorter calculation intervals are less erroneous. They further argued to use breath-based moving averages instead of time averages. In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003]. In a classic opinion piece, @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004. However, @robergs2010 acknowledged the lack of accessible software implementing such procedures.

@@

In this thesis I will review the usage and reporting of different data processing strategies in the scientific literature and investigate their influence on the V̇O~2max~.

To date, the practices of data processing to determine V̇O~2max~ in the actual scientific work remain largely unknown. Previous research on this topic has only performed unsystematic searches [@midgley2007] or non-representative surveys [@robergs2010]. Moreover, recent recommendation [@robergs2010] as well as advances in measurement devices and analysis software may have changed processing practices in the past 15 years.

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies extensively compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. @martin-rincon2019 based their comparisons on a data set of sedentary or recreational trained athletes using two different metabolic carts. Therefore motivation and measurement devices may have interacted with the influence of processing strategies. Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies. No research has yet compared a variety of systematically derived strategies among a group of well-trained individuals using a standardized measurement set-up.

This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature. Secondly, I will compare different data processing strategies in relation to the most common applied on a set of 72 standardized treadmill tests in well-trained individuals. The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals. The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting. The results build a basis for providing new recommendations for data processing and its reporting in determining the V̇O~2max.~

>>>>

Performance in endurance sports is limited by the human physiology. Athletes need to supply energy to the contracting muscles for locomotion, a process that happens mainly via the oxidative phosphorylation. A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster [@holloszy1984]. The highest activity of oxidative phosphorylation on a cellular level can be approximated by measuring the maximum oxygen uptake (V̇O~2max~) on a whole-body level.

V̇O~2max~ is defined as the highest rate a body can consume oxygen [@bassett2000]. Determined by measuring gas exchange data, it is one of the most common assessed exercise parameters in sports science and medicine. V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups [@reaburn2008; @costill1973; @tanaka1990] and can be regarded the most relevant physiological parameter for predicting endurance performance [@bassett2000]. Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~

Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion. In a laboratory setting such tests commonly take place on treadmills or cycling ergometer. Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value. Therefore most exercise protocols consist of a continuous or stepwise increase with an optimal during around 8-12 minutes [@buchfuhrer1983 ; @yoon2007]. While this narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008]. Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.

Early research on V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017]. Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001]. While mixing chambers are regarded as more reliable [@beijst2013], measuring breath-by-breath generates data with a higher temporal resolution [@roecker2005]. The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently.

Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades [@taylor1955; @howley1995; @poole2008; @poole2017; @green2018]. To identify a true maximum, researchers commonly evaluate a set of parameters measured during ramp tests --- the so called 'V̇O~2max~ criteria' [@howley1995]. If these criteria are (partly) not fulfilled, researchers are advised to speak of a peak oxygen uptake instead of V̇O~2max~ [@howley1995]. In this thesis I will not distinguish between peak and maximum oxygen uptake, as the criteria for V̇O~2max~ (e.g. the primary criterion of a plateau in oxygen uptake or the secondary criterion of the maximum respiratory quotient) do heavily rely on the data processing strategy used [@astorino2009]. Thus, I speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.

@@

Measured breath-by-breath data is noisy (see grey points in @fig-example). Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) lead to a highly fluctuating raw oxygen uptake. For interpretation the raw data requires some form of processing.

Different data processing strategies influence measured parameters of gas exchange. As soon as when the first automated systems for gas exchange measurement were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values. Since then, many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that shorter calculation intervals lead to higher V̇O~2max~ values [@johnson1998; @astorino2009; @sell2021]. @midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes. Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

```{r}
#| echo: false
#| fig-cap: "Example of raw breath-by-breath data processed by different strategies during the final minutes of a ramp test to exhaustion. Grey points display oxygen uptake from the single breaths. The moving averages were calculated over 30 breaths and seconds, respectively. The binned average was calculated over 30-second intervals with the mean values aligned to the center of each interval. For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied. For more details on the used data processing strategies, see the methods section. Data Source: [@quittmann_unpubl_b]"
#| fig-scap: "Example of raw breath-by-breath data processed by different strategies during the final minutes of a ramp test to exhaustion."
#| label: fig-example
#| out-width: '100%'

knitr::include_graphics("../plots/example.png")
```

@@

Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data, @robergs2003 opposed these conclusion, stating that shorter calculation intervals are less erroneous. They further argued to use breath-based moving averages instead of time averages. In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003]. In a classic opinion piece, @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004. However, @robergs2010 acknowledged the lack of accessible software implementing such procedures.

@@

In this thesis I will review the usage and reporting of different data processing strategies in the scientific literature and investigate their influence on the V̇O~2max~. To date, the practices of data processing to determine V̇O~2max~ in the actual scientific work remain largely unknown. Previous research on this topic has only performed unsystematic searches [@midgley2007] or non-representative surveys [@robergs2010]. Moreover, recent recommendation [@robergs2010] as well as advances in measurement devices and analysis software may have changed processing practices in the past 15 years.

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. @martin-rincon2019 provided formula for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts. Therefore motivation and measurement devices may have interacted with the influence of processing strategies.

Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies. No research has yet compared a variety of systematically derived strategies among a group of well-trained individuals using a standardized measurement set-up.

This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature. Secondly, I will compare different data processing strategies in relation to the most commonly applied on a set of 72 standardized treadmill tests in well-trained individuals. The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals. The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting. The results build a basis for providing new recommendations for data processing and its reporting for determining the V̇O~2max.~

### 2022-05-13 thesis/methods.qmd
Change layout

---
title: ''
format: pdf
bibliography: ../references.bib
---

@@

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker2020]. Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A.1). Major deviations will also be explicitly stated within the methods section. All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).

I conducted all analyses using R Version 4.1.2 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022]. The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a] . The attached packages and default settings are documented in Appendix A.4.

@@

The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.

The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A.2 contains the corresponding reporting checklist.

@@

| Source         | Search String                                                                                                                                                                                                            |
|----------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| PubMed         | `((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))` |
| Web of Science | `(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`                              |

: Search strings for the systematic scoping review. {#tbl-search}

The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI). Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step. After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles. @tbl-titlescan displays the exclusion terms for the automated title screening.

| Search terms for exclusion by title                                               |
|-----------------------------------------------------------------------------------|
| review, correction, meta-analysis, comment, retraction, editorial, erratum, reply |

: Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded. {#tbl-titlescan}

@@

The abstracts from the articles included in the random sample were blinded for scanning. This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix A.3 for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion. When the screeners disagreed in their assessments, they resolved the conflict by discussion.

|     | Criterium                                       | Details                                                                                                                                 |
|-----|-------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|
| A\* | not in English                                  | Full text only available in non-English language                                                                                        |
| B\* | no primary research                             | research was no original investigation or only a reanalysis of data                                                                     |
| C\* | research not in humans                          | research was conducted in animals                                                                                                       |
| D\* | V̇O~2max~ only estimated                         | V̇O~2max~ was only approximated by means of a predictive equation                                                                        |
| E   | no appropriate test protocol                    | Protocol for V̇O~2max~ testing did either not include exercise to voluntial exhaustion or was to long (\>20 min) for a reliable estimate |
| F   | no information regarding the exclusion criteria | crucial information on V̇O~2max~ testing that allowed the evaluation of the other exclusion criteria were missing                        |

: Exclusion criteria for the screening process. During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed. {#tbl-exclusion}

After the abstract screening I retrieved the full texts for all articles remaining in the review. The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria). Conflicts were resolved by discussion.

All data exclusion steps are documented in an Markdown script on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/review.md).

@@

Gas exchange data were recorded using a ZAN 600 device (nSpire Health, Inc., Longmont, CO, United States of America). The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement. The measured breath-by-breath data is available on [GitHub](https://github.com/smnnlt/vo2max-processing/tree/master/data/ramptests).

@@

#### Time based averaging

Moving time based averages were calculated by initially linearly interpolating the breath-by-breath data to seconds. Subsequently a (center aligned) moving average was calculated over a defined timespan. These processing steps are implemented in the spiro package [@nolte2022]. For calculating the moving average over 30 seconds for example, I used the functions `spiro(data) |> spiro_max(30)`.

Binned time averages were calculated using a custom function (available in the [analysis script](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md) on GitHub). Breath-by-breath data was initially interpolated to full seconds and then binned into consecutive interval of a constant length. The average of each interval was aligned to its center. Incomplete intervals (i.e. the last seconds of measurement) were not considered in the analysis. Note that some authors use a different procedure for determining their bins, starting by the end point of the measurement. However, defining bins beginning by the start of the measurement is a common output option for many gas exchange data analysis software.

#### Breath based averaging

@@

#### Butterworth filtering

Butterworth filters are a class of recursive digital filters. @robergs2010 and @weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data. However @robergs2010 missed to account for the time lag introduced by Butterworth filters. Therefore I applied a zero-phase Butterworth filter by means of a forward-backward filtering. While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it can non be used online (i.e. in real time). However, for the present application, an offline filter is sufficient. The forward-backward filtering also introduces transients at both ends of the signal [@gustafsson1996]. I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipi.signal.filtfilt()` function [@virtanen2020]). I used 3 as the order parameter and 0.04 as the cut-off frequency for each filter [@robergs2010]. Note that due to the forward-backward approach both the overall order and overall cut-off frequency are different from these values. The described approach is implemented in the spiro package [@nolte2022], and can be used by calling, for example, the functions `spiro(data) |> spiro_max("0.04fz3")`.

@@

To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level I calculated the median and 10%- and 90%-quantiles of each processing strategy.

>>>>

---
title: ''
format: pdf
crossref:
  sec-prefix: appendix
bibliography: ../references.bib
---

@@

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker2020]. Any deviations from the preregistration are indicated in the 'Transparent Changes' document (@sec-tchanges). Major deviations will also be explicitly stated within the methods section. All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).

I conducted all analyses using R Version 4.1.2 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022]. The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a]. The attached packages and default settings are documented in @sec-tech.

@@

The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans. The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018a]. @sec-checklist contains the corresponding reporting checklist.

@@

```{r}
#| echo: false
#| tbl-cap: "Search strings for the systematic scoping review."
#| label: tbl-search

search_df <- data.frame(
    s = c("PubMed", "Web of Science"),
    B = c('`((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`', '`(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`')
)

kableExtra::kbl(
  search_df,
  format = "latex",
  col.names = c("Source", "Search Strings"),
  booktabs = TRUE
) |> 
  kableExtra::column_spec(
    2, width = "12cm"
  )
```

The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI). Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step. After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles. All titles that contained one of the following words were excluded: review, correction, meta-analysis, comment, retraction, editorial, erratum, reply.

@@

The abstracts from the articles included in the random sample were blinded for scanning. This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see @sec-blindabstract for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion. When the screeners disagreed in their assessments, they resolved the conflict by discussion.

```{r}
#| echo: false
#| tbl-cap: "Exclusion criteria for the screening process."
#| label: tbl-exclusion

excl_df <- data.frame(
    id = c("A*", "B*", "C*", "D*", "E", "F"),
    crit = c(
      "not in English", "no primary research", "research not in humans", "VO2max only estimated",
      "no appropriate test protocol", "no information regarding the exclusion criteria"),
    info = c(
      "Full text only available in non-English language",
      "research was no original investigation or only a reanalysis of data", 
      "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", 
      "Protocol for VO2max testing did either not include exercise to voluntary exhaustion or was to long (>20 min) for a reliable estimate", 
      "no information regarding the exclusion criteria | crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing"
      )
)

out <- kableExtra::kbl(
  excl_df,
  format = "latex",
  col.names = c("", "Criterion", "Details"),
  booktabs = TRUE
) |> 
    kableExtra::column_spec(
    2, width = "4cm"
  ) |>
    kableExtra::column_spec(
    3, width = "9cm"
  ) |>
  kableExtra::footnote(
    general = "*During the abstract screening only the criteria marked with an asterix were evaluated",
    general_title = ""
  )

knitr::asis_output(stringr::str_replace_all(out, "VO2max", "${\\\\dot{\\\\textsf{V}}\\\\textsf{O}_{\\\\textsf{2max}}}$"))
```

After the abstract screening I retrieved the full-texts for all remaining articles. The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria). Conflicts were resolved by discussion between examiners. All data exclusion steps are documented in an Markdown script on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/review.md).

@@

Gas exchange data were recorded using a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America). The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement. The measured breath-by-breath data is available on [GitHub](https://github.com/smnnlt/vo2max-processing/tree/master/data/ramptests).

@@

Moving time based averages were calculated by first linearly interpolating the breath-by-breath data to seconds. Subsequently a (center aligned) moving average was calculated over a defined timespan. These processing steps are implemented in the spiro package [@nolte2022]. For calculating the moving average over 30 seconds for example, I used the functions `spiro(data) |> spiro_max(30)`.

Binned time averages were calculated using a custom function (available in the [analysis script](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md) on GitHub). Breath-by-breath data was initially interpolated to full seconds and then binned into consecutive interval of a constant length. The average of each interval was aligned to its center. Incomplete intervals (i.e. the last seconds of measurement) were not considered in the analysis. Note that some authors use a different procedure for determining their bins, starting by the end point of the measurement. However, defining bins beginning by the start of the measurement is a common output option for many gas exchange data analysis software (e.g. Cosmed Omnia).

@@

Butterworth filters are a class of recursive digital filters. @robergs2010 and @weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data. However @robergs2010 missed to account for the time lag introduced by Butterworth filters. Therefore I applied a zero-phase Butterworth filter by means of a forward-backward filtering. While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it cannot be used online (i.e. in real time). However, for the present application, an offline filter is sufficient. The forward-backward filtering also introduces transients at both ends of the signal [@gustafsson1996]. I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipy.signal.filtfilt()` function [@virtanen2020]). I used 3 as the order parameter and 0.04 as the cut-off frequency for each filter [@robergs2010]. Note that due to the forward-backward approach both the overall order and overall cut-off frequency are different from these values. The described approach is implemented in the spiro package [@nolte2022], and can be used by calling, for example, the functions `spiro(data) |> spiro_max("0.04fz3")`.

@@

To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level, I calculated the median and 10%- and 90%-quantiles of each processing strategy.
